---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

library(tidyverse)
library(deflateBR)
```

# deflateBR

<!-- badges: start -->
  [![R-CMD-check](https://github.com/meirelesff/deflateBR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/meirelesff/deflateBR/actions/workflows/R-CMD-check.yaml)
  <!-- badges: end -->

`deflateBR` is an `R` package used to deflate nominal Brazilian Reais using several popular price indexes.

## How does it work?

The `deflateBR`'s main function, `deflate`, requires three mandatory arguments to work: a `numeric` vector of nominal Reais (`nominal_values`); a `Date` vector of corresponding dates (`nominal_dates`); and a reference date (`real_date`), used to deflate the values. The reference date can be provided either as a character string in 'MM/YYYY' format or as a Date object. Additionally, you can specify the price index using the `index` argument and control output verbosity with the `verbose` argument. An example:

```{r}
# Load the package
library(deflateBR)

# Deflate January 2000 reais
deflate(nominal_values = 100, nominal_dates = as.Date("2000-01-01"), real_date = "01/2018")
```

Behind the scenes, `deflateBR` requests data from [IPEADATA](http://www.ipeadata.gov.br/)'s API on one these five Brazilian price indexes: [IPCA](https://ww2.ibge.gov.br/english/estatistica/indicadores/precos/inpc_ipca/defaultinpc.shtm) and [INPC](https://ww2.ibge.gov.br/english/estatistica/indicadores/precos/inpc_ipca/defaultinpc.shtm), maintained by [IBGE](https://ww2.ibge.gov.br/home/); and [IGP-M](http://portalibre.fgv.br/main.jsp?lumChannelId=402880811D8E34B9011D92B6160B0D7D), [IGP-DI](http://portalibre.fgv.br/main.jsp?lumChannelId=402880811D8E34B9011D92B6160B0D7D), and [IPC](http://portalibre.fgv.br/main.jsp?lumChannelId=402880811D8E34B9011D92B7350710C7) maintained by [FGV/IBRE](http://portalibre.fgv.br/main.jsp?lumChannelId=402880811D8E2C4C011D8E33F5700158). To select one of the available price indexes, just provide one of the following options to the `index =` argument: `ipca`, `igpm`, `igpdi`, `ipc`, and `inpc`. In the following, the INPC index is used.

```{r}
# Deflate January 2000 reais using the INPC price index
deflate(100, as.Date("2000-01-01"), "01/2018", index = "inpc")
```

The reference date can also be provided as a Date object for convenience:

```{r eval=FALSE}
# Same calculation using Date object for reference date
deflate(100, as.Date("2000-01-01"), as.Date("2018-01-01"), index = "inpc")
```

For silent operation without progress messages, you can use the `verbose` parameter:

```{r eval=FALSE}
# Silent operation - no progress messages or bars (using Date object)
deflate(100, as.Date("2000-01-01"), as.Date("2018-01-01"), index = "ipca", verbose = FALSE)
```

### Data Caching

Starting from version 1.2.0, `deflateBR` includes a caching system that stores downloaded price index data locally to improve performance. By default, caching is enabled and data is saved to a "cache" directory in your working directory:

```{r eval=FALSE}
# First call downloads data and saves to cache
deflate(100, as.Date("2000-01-01"), "01/2018", "ipca")
#> No cache found for IPCA index
#> Cache not available, downloading from IPEA API...
#> Saving IPCA data to cache...
#> Cached 547 records for IPCA index

# Subsequent calls load from cache (much faster)
deflate(200, as.Date("2001-01-01"), "01/2018", "ipca")
#> Found valid cache for IPCA index (547 records)
#> Loading IPCA data from cache...
#> Loaded 547 records from cache (1979-12 to 2025-06)
```

You can disable caching if you always want fresh data from the API:

```{r eval=FALSE}
# Disable caching - always download from API
deflate(100, as.Date("2000-01-01"), "01/2018", "ipca", cache = FALSE)
```

With the same syntax, a vector of nominal Reais can also be deflated, what is useful while working with `data.frames`:

```{r}
# Create some data
df <- tibble::tibble(reais = seq(101, 200),
                     dates = seq.Date(from = as.Date("2001-01-01"), by = "month", length.out = 100)
                     )

# Reference date to deflate the nominal values
reference <- "01/2018"

# Deflate using IGP-DI
head(deflate(df$reais, df$dates, reference, "igpdi"))
```

### Working with the tidyverse

For `tidyverse` users, the `deflate` function can be easily used inside code chains:

```{r}
library(tidyverse)

df %>%
  mutate(deflated_reais = deflate(reais, dates, reference, "ipca"))
```


## Installing

Install the latest released version from CRAN with:

```{r, eval=FALSE}
install.packages("deflateBR")
```

To install the development version of the package, use:
```{r, eval=FALSE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("meirelesff/deflateBR")
```

## Methodology

Following standard practice, seconded by the [Brazilian Central Bank](https://www3.bcb.gov.br/CALCIDADAO/publico/metodologiaCorrigirIndice.do?method=metodologiaCorrigirIndice), the `deflateBR` adjusts for inflation by multiplying nominal Reais by the ratio between the original and the reference price indexes. For example, to adjust 100 reais of January 2018, with IPCA index of 4916.46, to August 2018, with IPCA of 5056.56 in the previous month, we first calculate the ratio between the two indexes (e.g., 5056.56 / 4916.46 = 1.028496) and then multiply this value by 100 (e.g., 102.84 adjusted Reais). The `deflate` function gives exactly the same result:

```{r}
deflate(100, as.Date("2018-01-01"), "08/2018", "ipca")
```

## Citation

To cite `deflateBR` in publications, please use: 

```{r,eval=FALSE}
citation('deflateBR')
```


## Author

[Fernando Meireles](http://fmeireles.com)
